Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-08-30 TACO-paper-v2/coffee.tex
Add a few other lines, bye bye

\begin{figure}
\begin{alltt}
\scriptsize
\textbf{Input:} element matrix (2D array, initialized to 0)
       element coordinates (array)
       coefficient fields (array, e.g. velocity)
\textbf{Output:} element matrix (2D array)
- Compute Jacobian from coordinates
- Define basis functions
- Compute element matrix in an affine loop nest
\end{alltt}
\caption{Structure of a local assembly kernel}
\label{code:general-structure}
\end{figure}

@@

\begin{figure}[t]
\begin{alltt}
\scriptsize
\textbf{Input:} ast: kernel abstract syntax tree
       wrapper: function invoking the kernel
       isa: SIMD Instruction Set Architecture
\textbf{Output:} assembly code (C string)
\textbf{// Analyze ast and build optimization plan }
it\_space = analyze(ast)
\textbf{if not} it\_space
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \textbf{return} wrapper + ast.from\_ast\_to\_c()
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs)
\textbf{// Optimize ast based on plan}
ast.licm()
ast.padding()
ast.align\_data()
\textbf{if} plan.permute
  ast.permute\_assembly\_loops()
\textbf{if} plan.split\_factor
  ast.split(plan.split\_factor)
\textbf{if} plan.uaj\_factor
  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf)
  ast.op\_vect(uaj)
\textbf{return} wrapper + ast.from\_ast\_to\_c()
\end{alltt}
\caption{Outline of the COFFEE architecture.}
\label{algo:PyOP2Compiler}
\end{figure}

@@

\begin{equation*}
\scriptsize
\forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q}(f_K, g_K) 
\end{equation*}

@@

\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. Whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.  
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation is about resource-aware execution. It is indeed true that, in the context of COFFEE, it has been successfully applied to improve register pressure; however, the general idea is to re-schedule (re-associate) operations so as to maximize some parameters. Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel. An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlolcpc}. 
\item The quantities in the local assembly expression are known at code generation time. This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.

>>>>

\begin{figure}
\begin{alltt}
\scriptsize
\textbf{Input:} element matrix (2D array, initialized to 0), coordinates (array), coefficients (array, e.g. velocity)
\textbf{Output:} element matrix (2D array)
- Compute Jacobian from coordinates
- Define basis functions
- Compute element matrix in an affine loop nest
\end{alltt}
\caption{Structure of a local assembly kernel}
\label{code:general-structure}
\end{figure}

@@

\begin{figure}[t]
\begin{alltt}
\scriptsize
\textbf{Input:} ast (kernel abstract syntax tree), wrapper (function invoking the kernel), 
       isa (SIMD Instruction Set Architecture)
\textbf{Output:} assembly code (C string)
\textbf{// Analyze ast and build optimization plan }
it\_space = analyze(ast)
\textbf{if not} it\_space
  ast.apply\_inter\_kernel\_vectorization(wrapper, isa)
  \textbf{return} wrapper + ast.from\_ast\_to\_c()
plan = cost\_model(it\_space.n\_inner\_arrays, isa.n\_regs)
\textbf{// Optimize ast based on plan}
ast.licm()
ast.padding()
ast.align\_data()
\textbf{if} plan.permute
  ast.permute\_assembly\_loops()
\textbf{if} plan.split\_factor
  ast.split(plan.split\_factor)
\textbf{if} plan.uaj\_factor
  uaj = MIN(plan.uaj\_factor, it\_space.j.size/isa.vf)
  ast.op\_vect(uaj)
\textbf{return} wrapper + ast.from\_ast\_to\_c()
\end{alltt}
\caption{Outline of the COFFEE architecture.}
\label{algo:PyOP2Compiler}
\end{figure}

@@

\begin{equation}
\label{eq:assembly-model}
\scriptsize
\forall_{i, j} ~~~ A_{ij}^K = \sum_{q=1}^{n_1} \sum_{k=1}^{n_2} \alpha_{k, q}(a, b, c, ...) \beta_{q, i, j}(a, c, d, e, ...) \gamma_{q}(f_K, g_K) 
\end{equation}

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions in $\beta$ only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. The intuition is that whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.  
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation concerns resource-aware execution. It is indeed true that, in the context of COFFEE, expression splitting has been successfully applied to improve register pressure; however, the basic idea is to re-schedule (re-associate) operations so as to maximize some generic parameters, so it is far more general. It could be used, for example, as a starting point to perform kernel fission; that is, splitting a kernel into multiple parts, each part characterized by memory requirements less stringent than the original code (a variant of this idea for non-affine loops in unstructured mesh applications has been adopted in~\cite{carlolcpc}). In Equation~\ref{eq:assembly-model}, for instance, not only any of the functions $\alpha$, $\beta$ and $\gamma$ could be split (assuming they include associative operators), but also the evaluation of $\alpha$ in the summation over \texttt{k} could be completely extracted from the expression and evaluated in a separate kernel, allowing to save space for the allocation of $b$. 
\item The quantities in the local assembly expression are known at code generation time. This means that ``good'' code can be generated, for example with explicit loop bounds, statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
